---
title: "Notebook of COVID-19 analyses"
author: "Daniel Ricard (daniel.ricard@gmail.com)"
date: '`r paste0("Last modified timestap: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"))`'
output: html_notebook
---

# Context
Notes about analyses of COVID-19 data. Inspired by [an article on Towards Data Science](https://towardsdatascience.com/visualize-the-pandemic-with-r-covid-19-c3443de3b4e4). This document is an R markdown notebook available from my personal [viral2020 GitHub repository](https://github.com/ricardd/viral2020/).

# Worldwide cases and mortalities
Using information made available from the [European Centre for Disease Prevention and Control](https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide) and updated on a daily basis. There are likely other sources that are updated more regularly (e.g. the many sources used by [worldometers](https://www.worldometers.info/coronavirus/)), but this is a decent starting point.

```{r, include=FALSE}
## this is copied and pasted from the ECDC website

#these libraries need to be loaded
library(utils)
library(httr)

#download the dataset from the ECDC website to a local temporary file
GET("https://opendata.ecdc.europa.eu/covid19/casedistribution/csv", authenticate(":", ":", type="ntlm"), write_disk(tf <- tempfile(fileext = ".csv")))

#read the Dataset sheet into “R”. The dataset will be called "data".
data <- read.csv(tf, stringsAsFactors = FALSE)
names(data)[1] <- "dateRep"
## DR
data$dateRep <- as.Date(data$dateRep, format="%d/%m/%Y") ## DR, incoming data is formatted day/month/year and was not recognised correctly as a date
## compute the cumulative number of cases by country
## compute the cumulative number of fatalities by country
library(dplyr)
df <- data[order(data$countriesAndTerritories, data$dateRep),] %>% group_by(countriesAndTerritories) %>% mutate(cumsum.cases = cumsum(cases)) %>% mutate(cumsum.deaths = cumsum(deaths))
df$death.rate.percent <- 100*(df$cumsum.deaths / df$cumsum.cases)
df$cases.per.e06 <- (df$cumsum.cases / df$popData2018)*1E06
df$deaths.per.e06 <- (df$cumsum.deaths / df$popData2018)*1E06
## make sure that all countries have a 3-letter code so that they can be merged with the other data frame linking each country to a continent
#table(df$countryterritoryCode, exclude=FALSE)
#t.df <- df[df$countryterritoryCode=="",]
#table(t.df$countriesAndTerritories)

df[df$countriesAndTerritories=="Bonaire, Saint Eustatius and Saba","countryterritoryCode"] <- "BES"
df[df$countriesAndTerritories=="Falkland_Islands_(Malvinas)","countryterritoryCode"] <- "FLK"


```

```{r, include=FALSE}
## https://datahub.io/JohnSnowLabs/country-and-continent-codes-list
continent.df <- read.csv("country-and-continent-codes-list-csv_csv.txt")
#head(continent.df)
#df[which(df$countryterritoryCode==""),]
#table(continent.df$Country_Name)
#continent.df[which(continent.df$Country_Name=="Anguilla"),]
df.continent <- merge(df, continent.df, by.x="countryterritoryCode", by.y="Three_Letter_Country_Code")
```

```{r, echo=FALSE}
## a table showing today's number of cases in decreasing order
for.kable <- ungroup(df) %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) %>%
 arrange(desc(cumsum.cases))
```

Worldwide, there were `r format(sum(for.kable$cumsum.cases),big.mark=",")` confirmed cases and `r format(sum(for.kable$cumsum.deaths),big.mark=",")` deaths as this document was generated using the data obtained from the ECDC (HTML document generated `r format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")`). A total of `r nrow(for.kable)` countries had COVID-19 reported cases, `r nrow(for.kable[for.kable$cumsum.cases>100,])` countries had more than 100 reported cases,  and `r nrow(for.kable[for.kable$cumsum.deaths>100,])` countries had more than 100 reported deaths.

## Number of new deaths vs. cumulative deaths

Plot the number of new deaths or cases as a function of the cumulative number of deaths or cases and use a log scale on both axes. We do this for the number of deaths in the 10 countries with the most fatalities to date.

```{r, fig.height=6, fig.width=8, echo=FALSE}
library(dplyr)
#obtain top 20 and 10 country for fatalities
top20.deaths <- ungroup(df) %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) %>%
 top_n(n=20, wt=cumsum.deaths) %>%
 arrange(desc(cumsum.deaths))

top10.deaths <- ungroup(df) %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) %>%
 top_n(n=10, wt=cumsum.deaths) %>%
 arrange(desc(cumsum.deaths))

library(ggplot2)
require(ggrepel)

g <- 
  ggplot(filter(df, countriesAndTerritories %in% top10.deaths$countriesAndTerritories, deaths>0, cumsum.deaths>=10), aes(cumsum.deaths, deaths)) +
  geom_line(aes(group=countriesAndTerritories, color=countriesAndTerritories)) +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Cumulative number of deaths") +
  ylab("Daily number of deaths") +
  geom_text_repel(aes(label=countriesAndTerritories, color=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
theme_minimal(base_size=14) +
 theme(legend.position = "none") 
#+
#  geom_point(data=filter(df, countriesAndTerritories == "Italy", dateRep=format(Sys.time(),"%Y-%m-%d")))
g
```

Let's now do the same for another subset of countries.

```{r, fig.height=6, fig.width=8, echo=FALSE}
library(dplyr)
my.countries <- c("US","AT","SE","CZ","IT","ES","DE","FR","CA","AU","KR","NZ","DK","BR","EC")

g <- 
  ggplot(filter(df, geoId %in% my.countries, deaths>0, cumsum.deaths>=10), aes(cumsum.deaths, deaths)) +
  geom_line(aes(group=countriesAndTerritories, color=countriesAndTerritories)) +
  scale_x_log10() +
  scale_y_log10() +
  xlab("Cumulative number of deaths") +
  ylab("Daily number of deaths") +
  geom_text_repel(aes(label=countriesAndTerritories, color=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
theme_minimal(base_size=14) +
 theme(legend.position = "none") 
g
```


## Progression since 100 confirmed cases 
Present the number of cumulative cases starting on the date when 100 or more cases were recorded in each country, and use a logarithmic scale for the y axis. 

```{r, fig.height=6, fig.width=8, echo=FALSE}
first.day.100.cases <- aggregate(dateRep~countriesAndTerritories, data=df[which(df$cumsum.cases>=100),],min)
first.day.10.deaths <- aggregate(dateRep~countriesAndTerritories, data=df[which(df$cumsum.deaths>=10),],min)
df2 <- merge(merge(df, first.day.100.cases, by="countriesAndTerritories"), first.day.10.deaths, by="countriesAndTerritories")
names(df2)[c(2,17,18)] <- c("dateRep","date.first.100.cases","date.first.10.deaths")
df2$days.since.100.cases <- df2$dateRep - df2$date.first.100.cases
df2$days.since.10.deaths <- df2$dateRep - df2$date.first.10.deaths


n <- df2 %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) %>%
 top_n(20, cumsum.cases) %>%
 arrange(desc(cumsum.cases))

g<-
df2 %>%
  filter(countriesAndTerritories %in% n$countriesAndTerritories, as.numeric(df2$days.since.100.cases)>0) %>%
  ggplot(aes(as.numeric(days.since.100.cases), cumsum.cases, group=countriesAndTerritories, color=countriesAndTerritories)) +
 geom_line() +
 theme_minimal(base_size=14) +
  ylab("Cumulative confirmed number of cases") +
  ggtitle("Top 20 countries - cumulative") +
 geom_text_repel(aes(label=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
 theme_minimal(base_size=14) +
 theme(legend.position = "none")


#g
# and in log space
g + scale_y_log10() 

```

And do the same for the number of cumulative cases per one million people.

```{r, fig.height=6, fig.width=8, echo=FALSE}
g<-
df2 %>%
  filter(countriesAndTerritories %in% n$countriesAndTerritories, as.numeric(df2$days.since.100.cases)>0) %>%
  ggplot(aes(as.numeric(days.since.100.cases), cases.per.e06, group=countriesAndTerritories, color=countriesAndTerritories)) +
 geom_line() +
 theme_minimal(base_size=14) +
  ylab("Cumulative cases per million people") +
  ggtitle("Top 20 countries - cumulative") +
 geom_text_repel(aes(label=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
 theme_minimal(base_size=14) +
 theme(legend.position = "none")


#g
# and in log space
g + scale_y_log10() 

```

## Country-level summaries
How are different countries doing? I am testing a function that takes a country as an argument and fits both an exponential growth and a logistic growth model to the number of cumulative cases.

```{r, fig.height=6, fig.width=8, echo=FALSE}
my.logistic.fct <- function(a=100, b=0.1, c=1E03, ti){
  out <- c / (1 + (((c-a)/a)*exp(-b*ti)))
  return(out)
}


country.deaths.fct <- function(g.id){
  this.df <- filter(df2, geoId==g.id, cumsum.deaths>=10)

  ## 15 deaths per one million as a starting value for the asymptote c
  c.init <- 15 * (unique(this.df$popData2018)/1E06)
  #a.init <- c.init/100
    
  my.lm <- lm(log(cumsum.deaths)~days.since.10.deaths, data=this.df)
  pred.lm <- data.frame(dateRep=this.df$dateRep, cumsum.deaths=exp(fitted(my.lm)))

  initial.num.deaths <- this.df[as.numeric(this.df$days.since.10.deaths)==0,"cumsum.deaths"]
  
  my.logistic <- nls(cumsum.deaths ~ my.logistic.fct(a=initial.num.deaths, b, c, as.numeric(days.since.10.deaths)), data=this.df, start=list(b=0.34,c=c.init))
  logistic.pred <- data.frame(dateRep=this.df$dateRep, days.since.10.deaths=as.numeric(this.df$days.since.10.deaths), cumsum.deaths=fitted(my.logistic))
  ## inflection point
  t50 <- (log((coef(my.logistic)[2]-initial.num.deaths)/initial.num.deaths) - log(1))/coef(my.logistic)[1]
  n50 <- my.logistic.fct(a=initial.num.deaths, coef(my.logistic)[1], coef(my.logistic)[2], ti=t50)
  logistic.text <- paste0("Inflection point: ", round(t50,0), " days and ", round(n50,0), " deaths")
  
  country.name <- paste0(as.character(unique(this.df$countriesAndTerritories)), " - ", format(max(this.df$cumsum.deaths),big.mark=","), " cumulative deaths")
    
  
  g<-ggplot(this.df, aes(dateRep, cumsum.deaths)) +
    geom_line(aes(color="data")) +
    geom_point(aes(color="data")) +
    geom_line(data=pred.lm, aes(dateRep, cumsum.deaths, color="exponential model"), linetype=2) +
    geom_line(data=logistic.pred, aes(dateRep, cumsum.deaths, color="logistic model"), linetype=1) +
    guides(color=guide_legend("Model type")) +
    ggtitle(country.name) +
    geom_point(aes(x=min(dateRep)+t50, y=n50)) +
    annotate("text",x=min(this.df$dateRep)+(as.numeric(diff(range(this.df$dateRep)))*0.4), y=n50, label=logistic.text) +
    ylim(0,max(this.df$cumsum.deaths)*1.2)

  g
    
}

country.deaths.fct("IT")
country.deaths.fct("ES")
country.deaths.fct("US")
country.deaths.fct("FR")
country.deaths.fct("IR")
country.deaths.fct("UK")
country.deaths.fct("BE")
country.deaths.fct("CH")
country.deaths.fct("CA")

```

```{r, fig.height=6, fig.width=8, echo=FALSE}

country.cases.fct <- function(g.id){
  this.df <- filter(df2, geoId==g.id, cumsum.cases>=100)

  ## 1500 cases per one million as a starting value for the asymptote c
  c.init <- 1500 * (unique(this.df$popData2018)/1E06)
  #a.init <- c.init/100
    
  my.lm <- lm(log(cumsum.cases)~days.since.100.cases, data=this.df)
  pred.lm <- data.frame(dateRep=this.df$dateRep, cumsum.cases=exp(fitted(my.lm)))
  
  initial.num.cases <- this.df[as.numeric(this.df$days.since.100.cases)==0,"cumsum.cases"]
  my.logistic <- nls(cumsum.cases ~ my.logistic.fct(a=initial.num.cases, b,c,as.numeric(days.since.100.cases)), data=this.df, start=list(b=0.4,c=c.init))
  logistic.pred <- data.frame(dateRep=this.df$dateRep, days.since.100.cases=as.numeric(this.df$days.since.100.cases), cumsum.cases=fitted(my.logistic))
  ## inflection point
  t50 <- (log((coef(my.logistic)[2]-initial.num.cases)/initial.num.cases) - log(1))/coef(my.logistic)[1]
  n50 <- my.logistic.fct(a=initial.num.cases, coef(my.logistic)[1], coef(my.logistic)[2], ti=t50)
  logistic.text <- paste0("Inflection point: ", round(t50,0), " days and ", round(n50,0), " cases")
  
  country.name <- paste0(as.character(unique(this.df$countriesAndTerritories)), " - ", format(max(this.df$cumsum.cases),big.mark=","), " cumulative cases")
    
  
  g<-ggplot(this.df, aes(dateRep, cumsum.cases)) +
    geom_line(aes(color="data")) +
    geom_point(aes(color="data")) +
    geom_line(data=pred.lm, aes(dateRep, cumsum.cases, color="exponential model"), linetype=2) +
    geom_line(data=logistic.pred, aes(dateRep, cumsum.cases, color="logistic model"), linetype=1) +
    guides(color=guide_legend("Model type")) +
    ggtitle(country.name) +
    geom_point(aes(x=min(dateRep)+t50, y=n50)) +
    annotate("text",x=min(this.df$dateRep)+(as.numeric(diff(range(this.df$dateRep)))*0.4), y=n50, label=logistic.text) +
    ylim(0,max(this.df$cumsum.cases)*1.2)

  g
    
}


country.cases.fct("US")
country.cases.fct("IT")
country.cases.fct("ES")
country.cases.fct("CN")
country.cases.fct("DE")
country.cases.fct("FR")
country.cases.fct("IR")
country.cases.fct("UK")
country.cases.fct("CH")
country.cases.fct("TR")
country.cases.fct("BE")
country.cases.fct("NL")
country.cases.fct("AT")
country.cases.fct("KR")
country.cases.fct("CA")
country.cases.fct("PT")
country.cases.fct("BR")
country.cases.fct("IL")
country.cases.fct("AU")
country.cases.fct("NO")

#country.cases.fct("CZ")



#test.df <- data.frame(
#  ti=0:50)
#test.df$y <- my.logistic.fct(a=10, b=0.2, c=1E03, ti=test.df$ti)

#(log(100) - log(1))/0.2
#test.df[test.df$ti==23,]
```



