# Stale notes below

And do the same thing but using the number of confirmed cases instead.

```{r, fig.height=6, fig.width=8, echo=FALSE}
#obtain top 20 and 10 country for fatalities
top20.cases <- ungroup(df) %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) %>%
 top_n(n=20, wt=cumsum.cases) %>%
 arrange(desc(cumsum.cases))

top10.cases <- ungroup(df) %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) %>%
 top_n(n=10, wt=cumsum.cases) %>%
 arrange(desc(cumsum.cases))

g <- 
  ggplot(filter(df, countriesAndTerritories %in% top20.cases$countriesAndTerritories, cases>0), aes(cumsum.cases, cases)) +
  geom_line(aes(group=countriesAndTerritories, color=countriesAndTerritories)) +
  scale_x_log10() +
  scale_y_log10() +
    xlab("Cumulative number of cases") +
  ylab("Daily number of cases") +
  geom_text_repel(aes(label=countriesAndTerritories, color=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
theme_minimal(base_size=14) +
 theme(legend.position = "none") 

g
```


## Number of fatalities vs. time, plot for the top 20 countries based on number of fatalities

```{r fatalities, fig.height=6, fig.width=8, echo=FALSE}
#plot top 20

require(ggplot2)
require(ggrepel)


g <- ggplot(filter(df, countriesAndTerritories %in% top20.deaths$countriesAndTerritories, cases>0), 
 aes(dateRep, cumsum.deaths, color=countriesAndTerritories)) +
 geom_line() +
 geom_text_repel(aes(label=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
 theme_minimal(base_size=14) +
 theme(legend.position = "none") +
  ylab("Cumulative number of deaths") +
  ggtitle("Top 20 countries - cumulative number of deaths")
g
## and in log space
g + scale_y_log10()

```



## Number of confirmed cases vs. time, plot for the top 20 countries based on number of confirmed cases

```{r cases, fig.height=6, fig.width=8, echo=FALSE}
library(dplyr)
#obtain top 20 country
#plot top 20

require(ggplot2)
require(ggrepel)


g <- ggplot(filter(df, countriesAndTerritories %in% top20.cases$countriesAndTerritories, cases>0), 
 aes(dateRep, cumsum.cases, color=countriesAndTerritories)) +
 geom_line() +
 geom_text_repel(aes(label=countriesAndTerritories),
 function(d) d[d$dateRep == format(Sys.time(),"%Y-%m-%d"),]) +
 theme_minimal(base_size=14) +
 theme(legend.position = "none") +
  ylab("Cumulative confirmed number of cases") +
  ggtitle("Top 20 countries - cumulative number of cases")
g
## and in log space
g + scale_y_log10()

```



## Number of confirmed cases, as a table

```{r, results="asis", echo=FALSE}
library(knitr)
vars <- c("countriesAndTerritories", "cumsum.cases","cumsum.deaths", "popData2018", "cases.per.e06", "deaths.per.e06", "death.rate.percent"
)
kable(cbind(rank=1:nrow(for.kable),for.kable[,vars]), row.names = FALSE)
```





## A few countries plotted separately
Show the number of cumulative confirmed cases for a number of countries, sharing the same x axis from January 1st 2020 to the present.

```{r, fig.asp=5}

countries <- c("Japan","China","United States","Canada","Spain","Italy","France","Germany","Czechia","Norway","Australia","Argentina","Morocco","South Korea","Iran","Switzerland","Russia","Denmark","Singapore","Israel","United Kingdom","Chile")

this.data <- df[which(df$countriesAndTerritories %in% countries & df$cumsum.cases>100),]

ggplot(this.data,
 aes(dateRep, cumsum.cases)) +
 geom_line() + 
 scale_y_log10() + 
  facet_wrap(~countriesAndTerritories, ncol=2, scales="free_y")

```

## Fitting models
Fit exponential and logistic models to the data to see whether some countries are seeing a decrease in the number of cases. Let's take South Korea, the United States and Canada as examples to start.

```{r}
my.logistic.fct <- function(a=100, b=0.1, c=1E03, ti){
  out <- c / (1 + (((c-a)/a)*exp(-b*ti)))
  return(out)
}

#time <- 0:100
#plot(x=time,y=my.logistic(ti=time))


southkoreadata <- df2 %>%
  filter(countriesAndTerritories == "South_Korea" & cumsum.cases>100) %>%
  select(countriesAndTerritories,dateRep,cumsum.cases,cases,days.since.100.cases)

sk.lm <- lm(log(cumsum.cases)~dateRep, data=southkoreadata)
sk.lm.pred <- data.frame(dateRep=southkoreadata$dateRep, cumsum.cases=exp(fitted(sk.lm)))

sk.logistic <- nls(cumsum.cases ~ my.logistic.fct(a=100,b,c,as.numeric(days.since.100.cases)), data=southkoreadata, start=list(b=1.2,c=1E04))
sk.logistic.pred <- data.frame(dateRep=southkoreadata$dateRep, days.since.100.cases=as.numeric(southkoreadata$days.since.100.cases), cumsum.cases=fitted(sk.logistic))

g <- ggplot(southkoreadata, aes(dateRep, cumsum.cases)) +
  geom_line(aes(color="data")) +
  geom_point(aes(color="data")) +
  geom_line(data=sk.lm.pred, aes(dateRep, cumsum.cases, color="exponential model"), linetype=2) +
  geom_line(data=sk.logistic.pred, aes(dateRep, cumsum.cases, color="logistic model"), linetype=1) +
  guides(color=guide_legend("Model type")) +
  ggtitle("South Korea")
g

## what is the daily change in the number of new cases, once the number of daily cases is smoothed out to a 3-day moving average?
southkoreadata$cases.3day.avg <- NA
oo <- order(southkoreadata$dateRep)
southkoreadata <- southkoreadata[oo,] # order by date

days <- seq(from=range(southkoreadata$dateRep)[1]+2,to=range(southkoreadata$dateRep)[2],1)
for(d in days){
  l1 <- which(southkoreadata$dateRep==d)
  southkoreadata[l1,"cases.3day.avg"] <- mean(southkoreadata[l1:(l1-2),"cases"])
}

southkoreadata$R.3day.avg <- southkoreadata$cases.3day.avg / lag(southkoreadata$cases.3day.avg)

ggplot(southkoreadata, aes(dateRep, R.3day.avg)) +
  geom_line() + 
  geom_hline(yintercept = 1, lty=2) +
  ylim(0,1.8) +
  ggtitle("South Korea - R values by day, using a 3 day moving average")

## US
usdata <- df2 %>%
  filter(countriesAndTerritories == "United_States_of_America" & cumsum.cases>100) %>%
  select(countriesAndTerritories,dateRep,cumsum.cases,cases,days.since.100.cases)
us.lm <- lm(log(cumsum.cases)~dateRep, data=usdata)
us.lm.pred <- data.frame(dateRep=usdata$dateRep, cumsum.cases=exp(fitted(us.lm)))
#us.logistic 

us.logistic <- nls(cumsum.cases ~ my.logistic.fct(a=100,b,c,as.numeric(days.since.100.cases)), data=usdata, start=list(b=0.28,c=1E06))
us.logistic.pred <- data.frame(dateRep=usdata$dateRep, days.since.100.cases=as.numeric(usdata$days.since.100.cases), cumsum.cases=fitted(us.logistic))
long.df <- data.frame(days.since.100.cases=0:60)
us.logistic.pred.long <-data.frame(dateRep=seq(min(usdata$dateRep),min(usdata$dateRep)+60,1), days.since.100.cases=0:60, cumsum.cases=my.logistic.fct(a=100, b=coef(us.logistic)[1], c=coef(us.logistic)[2], ti=long.df$days.since.100.cases))

ggplot(usdata, aes(dateRep, cumsum.cases)) +
  geom_line(aes(color="data")) +
  geom_point(aes(color="data")) +
  geom_line(data=us.lm.pred, aes(dateRep, cumsum.cases, color="exponential model"), linetype=2) +
  geom_line(data=us.logistic.pred, aes(dateRep, cumsum.cases, color="logistic model - fit"), linetype=1) +
  geom_line(data=us.logistic.pred.long, aes(dateRep, cumsum.cases, color="logistic model - prediction"), linetype=2) +
  xlim(min(usdata$dateRep),min(usdata$dateRep)+60) +
  guides(color=guide_legend("Model type")) +
  ggtitle("United States") 


## what is the daily change in the number of new cases, once the number of daily cases is smoothed out to a 3-day moving average?
usdata$cases.3day.avg <- NA
oo <- order(usdata$dateRep)
usdata <- usdata[oo,] # order by date

days <- seq(from=range(usdata$dateRep)[1]+2,to=range(usdata$dateRep)[2],1)
for(d in days){
  l1 <- which(usdata$dateRep==d)
  usdata[l1,"cases.3day.avg"] <- mean(usdata[l1:(l1-2),"cases"])
}

usdata$R.3day.avg <- usdata$cases.3day.avg / lag(usdata$cases.3day.avg)

ggplot(usdata, aes(dateRep, R.3day.avg)) +
  geom_line() + 
  geom_hline(yintercept = 1, lty=2) +
  ylim(0,1.8) +
  ggtitle("United States - R values by day, using a 3 day moving average")


## Canada
canadadata <- df2 %>%
  filter(countriesAndTerritories == "Canada" & cumsum.cases>100) %>%
  select(countriesAndTerritories,dateRep,cumsum.cases,cases,days.since.100.cases)
canada.lm <- lm(log(cumsum.cases)~dateRep, data=canadadata)
canada.lm.pred <- data.frame(dateRep=canadadata$dateRep, cumsum.cases=exp(fitted(canada.lm)))

canada.logistic <- nls(cumsum.cases ~ my.logistic.fct(a=100,b,c,as.numeric(days.since.100.cases)), data=canadadata, start=list(b=0.3,c=1E04))
canada.logistic.pred <- data.frame(dateRep=canadadata$dateRep, days.since.100.cases=as.numeric(canadadata$days.since.100.cases), cumsum.cases=fitted(canada.logistic))

long.df <- data.frame(days.since.100.cases=0:60)
canada.logistic.pred.long <-data.frame(dateRep=seq(min(canadadata$dateRep),min(canadadata$dateRep)+60,1), days.since.100.cases=0:60, cumsum.cases=my.logistic.fct(a=100, b=coef(canada.logistic)[1], c=coef(canada.logistic)[2], ti=long.df$days.since.100.cases))

ggplot(canadadata, aes(dateRep, cumsum.cases)) +
  geom_line(aes(color="data")) +
  geom_point(aes(color="data")) +
  geom_line(data=canada.lm.pred, aes(dateRep, cumsum.cases, color="exponential model"), linetype=2) +
  geom_line(data=canada.logistic.pred, aes(dateRep, cumsum.cases, color="logistic model - fit"), linetype=1, lwd=1.2) +
  geom_line(data=canada.logistic.pred.long, aes(dateRep, cumsum.cases, color="logistic model - prediction"), linetype=2) +
  xlim(min(canadadata$dateRep),min(canadadata$dateRep)+60) +
  guides(color=guide_legend("Model type")) +
  ggtitle("Canada") 


## what is the daily change in the number of new cases, once the number of daily cases is smoothed out to a 3-day moving average?
canadadata$cases.3day.avg <- NA
oo <- order(canadadata$dateRep)
canadadata <- canadadata[oo,] # order by date

days <- seq(from=range(canadadata$dateRep)[1]+2,to=range(canadadata$dateRep)[2],1)
for(d in days){
  l1 <- which(canadadata$dateRep==d)
  canadadata[l1,"cases.3day.avg"] <- mean(canadadata[l1:(l1-2),"cases"])
}

canadadata$R.3day.avg <- canadadata$cases.3day.avg / lag(canadadata$cases.3day.avg)

ggplot(canadadata, aes(dateRep, R.3day.avg)) +
  geom_line() + 
  geom_hline(yintercept = 1, lty=2) +
  ylim(0,1.8) +
  ggtitle("Canada - R values by day, using a 3 day moving average")


## Italy
italydata <- df2 %>%
  filter(countriesAndTerritories == "Italy" & cumsum.cases>100) %>%
  select(countriesAndTerritories,dateRep,cumsum.cases,cases,days.since.100.cases)
italy.lm <- lm(log(cumsum.cases)~dateRep, data=italydata)
italy.lm.pred <- data.frame(dateRep=italydata$dateRep, cumsum.cases=exp(fitted(italy.lm)))

italy.logistic <- nls(cumsum.cases ~ my.logistic.fct(a=100,b,c,as.numeric(days.since.100.cases)), data=italydata, start=list(b=0.7,c=3E05))
italy.logistic.pred <- data.frame(dateRep=italydata$dateRep, days.since.100.cases=as.numeric(italydata$days.since.100.cases), cumsum.cases=fitted(italy.logistic))


ggplot(italydata, aes(dateRep, cumsum.cases)) +
  geom_line(aes(color="data")) +
  geom_point(aes(color="data")) +
  geom_line(data=italy.lm.pred, aes(dateRep, cumsum.cases, color="exponential model"), linetype=2) +
  geom_line(data=italy.logistic.pred, aes(dateRep, cumsum.cases, color="logistic model"), linetype=1) +
  guides(color=guide_legend("Model type")) +
  ggtitle("Italy") 


## what is the daily change in the number of new cases, once the number of daily cases is smoothed out to a 3-day moving average?
italydata$cases.3day.avg <- NA
oo <- order(italydata$dateRep)
italydata <- italydata[oo,] # order by date

days <- seq(from=range(italydata$dateRep)[1]+2,to=range(italydata$dateRep)[2],1)
for(d in days){
  l1 <- which(italydata$dateRep==d)
  italydata[l1,"cases.3day.avg"] <- mean(italydata[l1:(l1-2),"cases"])
}

italydata$R.3day.avg <- italydata$cases.3day.avg / lag(italydata$cases.3day.avg)

ggplot(italydata, aes(dateRep, R.3day.avg)) +
  geom_line() + 
  geom_hline(yintercept = 1, lty=2) +
  ylim(0,1.8) +
  ggtitle("Italy - R values by day, using a 3 day moving average")

```



## Number of cases per million people
Order by decreasing number of cases per one million people. The top of this list is composed of places with small populations.


```{r}
#df$cumsum.cases.per.e06 <- (df$cumsum.cases / df$popData2018) * 1E06
per.capita <- df %>% filter(dateRep == format(Sys.time(),"%Y-%m-%d")) 

ggplot(per.capita, aes(cases.per.e06)) +
  geom_histogram()

vars<-c("dateRep","countriesAndTerritories","cumsum.cases","popData2018","cases.per.e06")
kable(per.capita[order(per.capita$cases.per.e06,decreasing=TRUE),vars])

```

